Implementation of Set theory in \dedukti

### 2021-11-17 theoriezf/Article/Article.tex
Add files via upload

%\usepackage{paralist, tabularx}
\usepackage{oldlfont,amssymb,epsf,stmaryrd,epsfig,color}\usepackage{amsmath}
%\usepackage{hyperref}
%\usepackage{framed,multicol,changepage}
%\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
%\usepackage{skull,ifoddpage,marginnote}
%\usepackage{pifont}
%\usepackage{graphicx}
%\usepackage{ebproof}
%\usepackage{mathtools}
%\usepackage{bm}
%\usepackage{stmaryrd}
%\usepackage{cleveref}
%\usepackage{tikz}
%\usepackage{import} % For inkscape generated files

@@

\input{macros.tex}
\input{lambdapi.tex}

@@

Introduction : papier précédent, explications de la structure des graphes pointés, apparition de Dedukti depuis (passage de Deduction modulo à Lambpdapi calcul modulo), théorie U (arrow, omicron) \\

The language of pointed graphs : types, signature \\

The theory of pointed graphs : règles de réécritures \\

@@

papier + dedukti + bout de théorie U (arrow, omicron)

@@

\begin{lstlisting}
symbol eta : Graph → Node → Node → Prop;
symbol root : Graph → Node;
symbol cr : Graph → Node → Graph;
constant symbol o : Node;
constant symbol O : Node;
symbol i : Node → Node;
symbol i' : Node → Node;
symbol j : Node → Node;
symbol j' : Node → Node;
symbol I : Node → Prop;
symbol J : Node → Prop;
symbol ρ : Graph → Node;
symbol ρ^ : Node → Graph;
symbol Succ : Node → Node;
symbol Pred : Node → Node;
symbol Null : Node → Prop;
symbol Nat : Node → Prop;
symbol < : Node → Node → Prop;
symbol simeq : Graph → Graph → Prop;
symbol ∈ : Graph → Graph → Prop;
symbol join : Graph → Graph;
symbol pair : Graph → Graph → Graph;
symbol powerset : Graph → Graph;
symbol comp : Graph → El valuation → El formula → Graph;
symbol omega : Graph;
symbol Cl : Graph → Graph;
\end{lstlisting}

\section{The language of formulas}
\subsection{Formulas}

$$comp : \Graph \ra \El \ valuation \ra \El \ formula \ra \Graph$$

In $comp$ we may have noticed the presence of valuation and formula. 
Indeed, the lemmas in which the $comp$ symbol is used are only valid for a subset of formulas. The formulas need to have all of its quantifiers of sort $Graph$ and can only use the language $\in$, $\simeq$ and the classic logical connectives.

Thus we need to introduce a set to caraterize the validity domain of such lemmas. In order to achieve this goal, we define the constant $formula$ of type $\Set$ and the logical connectives related to this constant.

@@

@@

The next step is to interpret an object of type $formula$ into $Prop$. We introduce the constant $interpretation$ which receives a valuation of type $\El \ valuation$ and a formula of type $\El \ formula$ and return a $\Prop$.

@@

The interpreation of formulas obeys the following rewriting rules :

$$interpretation \ \sigma \ (eqF \ x \ y) \lra (\sigma \ x) \ \simeq \ (\sigma \ y)$$
$$interpretation \ \sigma \ (inF \ x \ y) \lra (\sigma \ x) \ \in \ (\sigma \ y)$$
$$interpretation \ \sigma \ (andF \ f \ g) \lra (interpretation \ \sigma \ f) \conj (interpretation \ \sigma \ g)$$
$$interpretation \ \sigma \ (orF \ f \ g) \lra (interpretation \ \sigma \ f) \disj  (interpretation \sigma \ g)$$
$$interpretation \ \sigma \ (impF \ f \ g) \lra (interpretation \ \sigma \ f) \imp (interpretation \ \sigma \ g)$$
$$interpretation \ \sigma \ fF \lra \blbot$$
$$interpretation \ \sigma \ tF \lra \bltop$$

We need to have a tool to update a valuation when we assign a variable. To do so, we introduce the constant $update$ of type $(\El \ nat \ra Graph) \ra \El \ nat \ra Graph \ra (\El \ nat \ra Graph)$ which takes as arguments a valuation, a natural number and a graph and returns a new valuation.

We define the symbol $update1$ of type $(\El \ nat \ \ra \ Graph) \ \ra \ \El \ nat \ \ra \ Graph \ \ra \ \El \ nat \ \ra \ (\El \ nat \ \ra \ Graph)$ which obeys the following rewriting rules :

$$update \ \sigma \ x \ a \ y \ \lra \ update1 \ \sigma \ x \ a \ y \ y$$
$$update \ \sigma \ zero \ a \ zero \ z \ \lra \ a$$
$$update \ \sigma \ zero \ a \ (s \ y) \ z \ \lra \ \sigma \ z$$
$$update \ \sigma \ (s \ x) \ a \ zero \ z \ \lra \ \sigma \ z$$
$$update \ \sigma \ (s \ x) \ a \ (s \ y) \ z \ \lra \ update1 \ \sigma \ x \ a \ y \ z$$

+ A finir avec update1
+ supprimer El valuation ou pas ?

@@

@@

\begin{center}
\begin{tabular}{|c|c||c|c|}
\hline Lemma & Number of lines in the proof & Lemma & Number of lines in the proof \\
\hline 3 & ? & 29 & ? \\
\hline 4 & ? & 30 & ? \\
\hline 5 & ? & 31 & ? \\
\hline 6 & ? & 32 & ? \\
\hline 7 & ? & 33 & ? \\
\hline 8 & ? & 34 & ? \\
\hline 9 & ? & 35 & ? \\
\hline 10 & ? & 36 & ? \\
\hline 11 & ? & 37 & ? \\
\hline 12 & ? & 38 & ? \\
\hline 13 & ? & 39 & ? \\
\hline 14 & ? & 40 & ? \\
\hline 15 & ? & 41 & ? \\
\hline 16 & ? & 42 & ? \\
\hline 17 & ? & 43 & ? \\
\hline 18 & ? & 44 & ? \\
\hline 19 & ? & 45 & ? \\
\hline 20 & ? & 46 & ? \\
\hline 21 & ? & 47 & ? \\
\hline 22 & ? & 48 & ? \\
\hline 23 & ? & 49 & ? \\
\hline 24 & ? & 50 & ? \\
\hline 25 & ? & 51 & ? \\
\hline 26 & ? & 52 & ? \\
\hline 27 & ? & 53 & ? \\
\hline 28 & ? & & \\
\hline
\end{tabular}
\end{center}

\section*{References}

>>>>

%\usepackage{paralist, tabularx}
\usepackage{oldlfont,amssymb,epsf,stmaryrd,epsfig,color}\usepackage{amsmath}
%\usepackage{hyperref}
%\usepackage{framed,multicol,changepage}
%\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
%\usepackage{skull,ifoddpage,marginnote}
%\usepackage{pifont}
%\usepackage{graphicx}
%\usepackage{ebproof}
%\usepackage{mathtools}
%\usepackage{bm}
%\usepackage{stmaryrd}
%\usepackage{cleveref}
%\usepackage{tikz}
%\usepackage{import} % For inkscape generated files
\usepackage{biblatex}
\usepackage{url}

@@

\input{macros.tex}
\input{lambdapi.tex}
\addbibresource{ref.bib}

@@

Introduction : introduction du papier précédent, apparition de Dedukti depuis (passage de Deduction modulo à Lambpdapi calcul modulo), théorie U (arrow, omicron) \\

The theory of pointed graphs :  explications détaillées de la structure des graphes pointés \\

The language of pointed graphs : types, signature, règles de réécritures \\

@@

@@

\begin{lstlisting}
symbol eta : Graph → Node → Node → Prop;
symbol root : Graph → Node;
symbol cr : Graph → Node → Graph;
constant symbol o : Node;
constant symbol O : Node;
symbol i : Node → Node;
symbol i' : Node → Node;
symbol j : Node → Node;
symbol j' : Node → Node;
symbol I : Node → Prop;
symbol J : Node → Prop;
symbol ρ : Graph → Node;
symbol ρ^ : Node → Graph;
symbol Succ : Node → Node;
symbol Pred : Node → Node;
symbol Null : Node → Prop;
symbol Nat : Node → Prop;
symbol < : Node → Node → Prop;
symbol simeq : Graph → Graph → Prop;
symbol ∈ : Graph → Graph → Prop;
symbol join : Graph → Graph;
symbol pair : Graph → Graph → Graph;
symbol powerset : Graph → Graph;
symbol comp : Graph → (El nat → Graph) → El formula → Graph;
symbol omega : Graph;
symbol Cl : Graph → Graph;
\end{lstlisting}

\subsection{Rewriting rules}

\section{The language of formulas}

We may have notice that the lemmas in which the $comp$ symbol is used are only valid for a subset of formulas. The formulas need to have all of its quantifiers of sort $Graph$ and can only use the language $\in$, $\simeq$ and the classic logical connectives.

Thus we need to introduce a set to caraterize the validity domain of such lemmas.

\subsection{Formulas}

In order to achieve this goal, we define the constant $formula$ of type $\Set$ and the logical connectives related to this constant.

@@

Henceforth, we are able to define in \dedukti
$$comp : \Graph \ra (\El \ nat \ \ra \ Graph) \  \ra \El \ formula \ra \Graph$$

@@

The next step is to interpret an object of type $formula$ into $Prop$. We introduce the constant $interpretation$ which receives a valuation of type $\El \ nat \ \ra \ Graph$ and a formula of type $\El \ formula$ and return a $\Prop$.

@@

We need to have a tool to update a valuation when we assign a variable. To do so, we introduce the constant $update$ of type $(\El \ nat \ra Graph) \ra \El \ nat \ra Graph \ra (\El \ nat \ra Graph)$ which takes as arguments a valuation, a natural number and a graph and returns a new valuation: $(update \ \sigma \ x \ a)$ is a new valuation that substitues $x$ by $a$ and acts like $\sigma$ for the other natural numbers.

To write a rewriting rule upon $update$, we need to be able to check if we apply $(update \ \sigma \ x \ a)$ to $z = x$ or to $z \neq x$.

We define the symbol $update1$ of type $(\El \ nat \ \ra \ Graph) \ \ra \ \El \ nat \ \ra \ Graph \ \ra \ \El \ nat \ \ra \ (\El \ nat \ \ra \ Graph)$. We keep in memory the argument of the valuation with the rewriting rule:

$$update \ \sigma \ x \ a \ y \ \lra \ update1 \ \sigma \ x \ a \ y \ y$$

We have two natural numbers to compare: $x$, which is substitued by $a$ by the $update$, and $y$ that is the argument we apply to the valuation. The technique we use to compare $x$ and $y$ is the following: we keep in memory $y$ in the variable $z$, we decrement the two numbers until either one or both are equal to zero. If both are equal to zero, then $x$ and $y$ are equal and we return $a$. If only one equals zero, then they are different and we return $\sigma \ z$.

$$update1 \ \sigma \ (s \ x) \ a \ (s \ y) \ z \ \lra \ update1 \ \sigma \ x \ a \ y \ z$$
$$ update1 \ \sigma \ zero \ a \ zero \ z \ \lra \ a$$
$$update1 \ \sigma \ zero \ a \ (s \ y) \ z \ \lra \ \sigma \ z$$
$$update1 \ \sigma \ (s \ x) \ a \ zero \ z \ \lra \ \sigma \ z$$

Now we have all the tools to define the rewiting rules of the interpreation of formulas:

$$interpretation \ \sigma \ (eqF \ x \ y) \lra (\sigma \ x) \ \simeq \ (\sigma \ y)$$
$$interpretation \ \sigma \ (inF \ x \ y) \lra (\sigma \ x) \ \in \ (\sigma \ y)$$
$$interpretation \ \sigma \ (andF \ f \ g) \lra (interpretation \ \sigma \ f) \conj (interpretation \ \sigma \ g)$$
$$interpretation \ \sigma \ (orF \ f \ g) \lra (interpretation \ \sigma \ f) \disj  (interpretation \sigma \ g)$$
$$interpretation \ \sigma \ (impF \ f \ g) \lra (interpretation \ \sigma \ f) \imp (interpretation \ \sigma \ g)$$
$$interpretation \ \sigma \ (allF \ x \ f) \ \lra \ \fa \ a, interpretation \ (update \ \sigma \ x \ a) \ f$$
$$interpretation \ \sigma \ (exF \ x \ f) \ \lra \ \ex \ a, interpretation \ (update \ \sigma \ x \ a) \ f$$
$$interpretation \ \sigma \ fF \lra \blbot$$
$$interpretation \ \sigma \ tF \lra \bltop$$

@@

The first lemma $x=x$ does not need to be implemented since it is already part of the standard library of \dedukti \ under the name $refl$ (which is polymorphic).

The second lemma is not required as it is a consequence of the rewriting rule of the polumorphic $=$ implemented in the stardard library of \dedukti: 
\begin{lstlisting}
constant symbol = {s} : El s → El s → Prop;
notation = infix 4;
rule π (@= $s $x $y) ↪ Π (P : El $s → Prop), π(P $x) → π(P $y);
\end{lstlisting}

\subsection{Useful mechanisms}

All the other lemmas except the ones where $comp$ is involved are proved using the blueprint in \cite{zermodulo53}. The complete proofs can be found in \url{https://github.com/ttraversie/zf/tree/main/theoriezf}.

\subsection{Some interesting proofs}

Lemma32

\subsection{Weak extensionnality}

We want to deduce $weak \ extensionality$ from $strong \ extensionality$ (i.e. lemma 41).

\begin{lstlisting}
opaque symbol lemma41 : Π (x y : El nat), Π (f : El formula), Π (c d : Graph), π(`∀ σ : (El nat → Graph), ((interpretation (update (update σ x c) y d) f)
∧ (`∀ a, `∀ a', `∀ b, ((a' ∈ a) ∧ (interpretation (update (update σ x a) y b) f)) ⇒ (`∃ b', ((b' ∈ b) ∧ (interpretation (update (update σ x a') y b') f))))
∧ (`∀ b, `∀ b', `∀ a, ((b' ∈ b) ∧ (interpretation (update (update σ x a) y b) f)) ⇒ (`∃ a', ((a' ∈ a) ∧ (interpretation (update (update σ x a') y b') f))))) ⇒ (c simeq d))

opaque symbol lemmaExt : Π (c d : Graph), π((`∀ x, (x ∈ c) ⇔ (x ∈ d)) ⇒ (c simeq d))
\end{lstlisting}

To do this, we need to prove the following intermediry lemma:

\begin{lstlisting}
opaque symbol lemmaHypExt : Π (c d : Graph), π((`∀ z, (z ∈ c) ⇔ (z ∈ d)) ⇒ ((((c simeq c) ∧ (d simeq d)) ∨ (c simeq d))
∧ (`∀ a, `∀ a', `∀ b, ((a' ∈ a) ∧ (((a simeq c) ∧ (b simeq d)) ∨ (a simeq b))) ⇒ (`∃ b', ((b' ∈ b) ∧ (((a' simeq c) ∧ (b' simeq d)) ∨ (a' simeq b')))))
∧ (`∀ b, `∀ b', `∀ a, ((b' ∈ b) ∧ (((a simeq c) ∧ (b simeq d)) ∨ (a simeq b))) ⇒ (`∃ a', ((a' ∈ a) ∧ (((a' simeq c) ∧ (b' simeq d)) ∨ (a' simeq b')))))))
\end{lstlisting}

We thus can prove $weak \ extensionality$ thanks to:

\begin{lstlisting}
opaque symbol lemmaExt : Π (c d : Graph), π((`∀ x, (x ∈ c) ⇔ (x ∈ d)) ⇒ (c simeq d))
≔ begin
assume c d H;
refine lemma41 zero one (orF (andF (eqF zero two) (eqF one three)) (eqF zero one)) c d (update (update (λ _, empty_set) two c) three d) (lemmaHypExt c d H);
end;
\end{lstlisting}

@@

\begin{center}
\begin{tabular}{|c|c||c|c|}
\hline Lemma & Number of lines in the proof & Lemma & Number of lines in the proof \\
\hline 3 & ? & 29 & ? \\
\hline 4 & ? & 30 & ? \\
\hline 5 & ? & 31 & ? \\
\hline 6 & ? & 32 & ? \\
\hline 7 & ? & 33 & ? \\
\hline 8 & ? & 34 & ? \\
\hline 9 & ? & 35 & ? \\
\hline 10 & ? & 36 & ? \\
\hline 11 & ? & 37 & ? \\
\hline 12 & ? & 38 & ? \\
\hline 13 & ? & 39 & ? \\
\hline 14 & ? & 40 & ? \\
\hline 15 & ? & 41 & ? \\
\hline 16 & ? & 42 & ? \\
\hline 17 & ? & 43 & ? \\
\hline 18 & ? & 44 & ? \\
\hline 19 & ? & 45 & ? \\
\hline 20 & ? & 46 & ? \\
\hline 21 & ? & 47 & ? \\
\hline 22 & ? & 48 & ? \\
\hline 23 & ? & 49 & ? \\
\hline 24 & ? & 50 & ? \\
\hline 25 & ? & 51 & ? \\
\hline 26 & ? & 52 & ? \\
\hline 27 & ? & 53 & ? \\
\hline 28 & ? & Weak extensionality & ? \\
\hline
\end{tabular}
\end{center}

\printbibliography
